Megan's code for Isoviz

### 2023-03-28 temp_code/MS_isoviz_code.Rmd
MS add plot junction to isoform

```{r}
# generate leafcutter cluster files and store for standard cell types
# this is publicly available A375 data used to design the screen

# get leafcutter data in proper format
clusters <- read_tsv("~/isoviz/data-raw/SRR6515912_out-regtools_.junc.a375_m30_p0.01_.sorted", col_names = FALSE, skip = 1) # 15,901
clusters %<>% 
  separate(X1, into = c("coords", "counts"), sep = " ") %>% 
  separate(counts, into = c("junc.counts", "cluster.counts"), sep = "/", convert = TRUE) %>%
  separate(coords, into = c("chr", "start", "end", "name"), sep = ":", convert = TRUE) %>% 
  separate(name, into = c("same", "cluster.n", "extra"), sep = "_") %>% 
  select(-same, -extra)

# remove clusters with no counts, not sure why these are here to begin with- only 7
# filter out chrM and chrY clusters
clusters %<>% filter(cluster.counts > 0, chr != "chrY", chr != "chrM") # 15,886

leafcutter_clusters = as.data.table(clusters)

save(leafcutter_clusters, file = "~/isoviz/data/leafcutter_clusters.rda")
```

```{r}
# could make this a function
# create gene name conversion file from gtf
gtf = read_tsv("~/isoviz/data-raw/gencode.v41.basic.annotation.gtf", col_names = FALSE, skip = 5) %>% filter(X3 == "transcript")
gtf$transcript_id = sapply(strsplit(as.character(gtf$X9), ";"), function(x) x[[2]])
gtf$transcript_id = sapply(gtf$transcript_id, gsub, pattern = "transcript_id", replacement = "")
gtf$transcript_id = sapply(as.character(gtf$transcript_id), gsub, pattern = '"', replacement = '')
gtf$transcript_id = sapply(as.character(gtf$transcript_id), gsub, pattern = ' ', replacement = '')

@@

gtf %>% dplyr::select(gene_id, trans_id = transcript_id, gene_name, transcript_name, gene_type, transcript_type) %>% distinct() %>%
  write_tsv("~/isoviz/data-raw/gencode_v41_gene-transcript-convert.txt", col_names = TRUE)

@@

```{r}
# modify psl file data to include other gene and transcript names
# currently using gencode basic anntotation file
convert = read_tsv("~/isoviz/data-raw/gencode_v41_gene-transcript-convert.txt", col_names = TRUE)

@@

```{r}
# make psl with intron coords instead of exon
# adapted from genomedata.R

mydataset <- fread("~/isoviz/data-raw/gencode.v41.basic.annotation.psl")

@@

for(id in transcript_ids){
  data = mydataset %>% filter(trans_id == id)
  new_ends = data$blockstarts[-1] + 1
  len = nrow(data)
  new_starts = data$blockends[-len]
  n = len-1
  trans_id = c(trans_id, rep(id, n))
  intron_starts = c(intron_starts, new_starts)
  intron_ends = c(intron_ends, new_ends)
}

@@

save(iso_intron_data, file = "~/isoviz/data/iso_intron_data.rda")

@@

```{r}
load(file = "~/isoviz/data/iso_intron_data.rda")
load(file = "~/isoviz/data/leafcutter_clusters.rda")

@@

  # filter for gene, for now, also filter for protein coding transcripts- can set this as a parameter later
  gene_iso_data = iso_intron_data %>% filter(grepl(gene, gene_id), transcript_type == "protein_coding")

  # join iso data with leafcutter data and mark any unannotated junctions
  gene_cluster = gene_iso_data %>% left_join(leafcutter_clusters, by = c("chr", "intron_starts" = "start", "intron_ends" = "end")) %>% filter(!is.na(cluster.n))
  gene = unique(gene_cluster$gene_id)
  name = unique(gene_cluster$gene_name)
  gene_strand = unique(gene_cluster$strand)
  cluster_ids = unique(gene_cluster$cluster.n)
  output = leafcutter_clusters %>% filter(cluster.n %in% cluster_ids) %>% group_by(cluster.n) %>% mutate(junc.per.cluster = n()) %>% 
    left_join(gene_iso_data, by = c("chr", "start" = "intron_starts", "end" = "intron_ends")) %>% 
    mutate(junc.usage = round((junc.counts/cluster.counts)*100, digits = 0), 
           transcript_name = ifelse(is.na(transcript_name), "Unannotated", transcript_name),
           trans_id = ifelse(is.na(trans_id), "Unannotated", trans_id),
           gene_id = ifelse(is.na(gene_id), gene, gene_id),
           gene_name = ifelse(is.na(gene_name), name, gene_name),
           strand = ifelse(is.na(strand), gene_strand, strand))

@@

to_plot = map_junction("ENSG00000001167.15") # NFYA
to_plot = map_junction("ENSG00000100320") # RBFOX2

@@

```{r}
load(file='~/isoviz/data/iso_data_complete.rda')
to_plot

# plot isoform plus leafcutter clusters
plot_junction_to_isoform = function(to_plot){

  # get info
  gene = unique(to_plot$gene_id)
  name = unique(to_plot$gene_name)
  transcripts = unique(to_plot$transcript_name)

  # I am copying the plot_isoforms function here, figure out how to integrate these later
  gene_trans =which(str_detect(iso_data_complete$gene_id, gene))
  gene_data = iso_data_complete[gene_trans,]

@@

  to_plot_ordered = gene_data %>% full_join(order_plot, by = c("transcript_name")) %>%
    arrange(trans_order, blockstarts) %>%
    mutate(y1 = trans_order - 0.2, y2 = trans_order + 0.2)
  to_plot_ordered = to_plot_ordered %>% mutate(seg_end = end)

  # make isoform plot
  p1 = ggplot() +
    geom_segment(aes(x = to_plot_ordered$start, y = to_plot_ordered$trans_order,
                     xend = to_plot_ordered$seg_end, yend = to_plot_ordered$trans_order)) +
    geom_rect(data = to_plot_ordered, mapping = aes(xmin = blockstarts, xmax = blockends, ymin = y1, ymax = y2)) +
    xlim(min_start, max_end) + ylim(0,y_max) +
    geom_text(aes(x = Inf, y = to_plot_ordered$trans_order, hjust = 0, label = to_plot_ordered$transcript_name)) +
    theme_bw() + ylab("") + xlab("") +
    theme(axis.text.y=element_blank(), axis.ticks.y=element_blank(), legend.position = "top") +
    ggtitle(paste0(name, " (", to_plot_ordered$strand[1], ")")) +
    theme(axis.text = element_text(size = 10), plot.title = element_text(hjust = 0.5)) + coord_cartesian(xlim = c(min_start, max_end))
    #coord_cartesian(clip = 'off')
    #theme(plot.margin = margin(0.5,1.5,0.5,0.5, "in"))

  # new code to plot leafcutter cluster info underneath
  clusters = unique(to_plot$cluster.n)
  introns = to_plot %>% select(-gene_id, -trans_id, -gene_name, -transcript_name) %>% distinct() %>% group_by(cluster.n) %>% mutate(junc_order = row_number())
  p2 = ggplot() + geom_rect(data = introns, mapping = aes(xmin = start, xmax = end, ymin = junc_order - 0.2, ymax = junc_order + 0.2)) + 
    xlim(min_start, max_end) + theme_bw() + facet_wrap(~ cluster.n, scales = "free_y", ncol = 1, strip.position = "top") + 
    xlab("Hg38 Genomic Position (bp)") + ylab("") + theme(axis.text.y=element_blank(), axis.ticks.y=element_blank()) +
    coord_cartesian(xlim = c(min_start, max_end))

  plot_grid(p1, p2, ncol = 1, align = "v", rel_heights = c(3, 3))
  ggsave("~/isoviz/plots/rbfox2_isoform_draft.pdf", width = 8, height = 4)

    # try using pivot for aggregating isoform

}
```

>>>>

Raw data -> Data for package
Leafcutter clusters
Run this on publicly available data for major cell types
But have function for people to read in their on bam file

Defining Intron clustering with Leafcutter
make a text file with the name of all bed files

# more stringent- better for calling unique clusters
python ~/Downloads/leafcutter/clustering/leafcutter_cluster_regtools.py -j a375_screen_cells_for_lfc.txt -m 30 -p 0.1 -o a375_screen_cells -l 500000 --includeconst --checkchrom

# less stringent (-m 5) which is better for detecting all singletons
python ~/Downloads/leafcutter/clustering/leafcutter_cluster_regtools.py -j a375_screen_cells_for_lfc.txt -m 5 -p 0.1 -o a375_screen_cells_5reads -l 500000 --includeconst --checkchrom

output is a text file
-p MINCLURATIO, --mincluratio=MINCLURATIO
                        minimum fraction of reads in a cluster that support a
                        junction (default 0.001)

--includeconst
```{r}
# This is data from our A375 screen cells and is the most accurate for our work
# now this includes constitutive exons which have a single junction per cluster
# required 30 total reads per cluster and junction to be 10% of total
# total reads in a375-cyt_Aligned.q20.sort.bam: 43,407,715
clusters <- read_tsv("~/isoviz/data-raw/a375_screen_cells_perind.constcounts.gz", col_names = FALSE, skip = 1) # 59,783
clusters %<>% 
  separate(X1, into = c("coords", "counts"), sep = " ") %>% 
  separate(counts, into = c("junc.counts", "cluster.counts"), sep = "/", convert = TRUE) %>%
  separate(coords, into = c("chr", "start", "end", "name"), sep = ":", convert = TRUE) %>% 
  separate(name, into = c("same", "cluster.n", "extra"), sep = "_") %>% 
  select(-same, -extra)

# remove clusters with no counts
# filter out chrM and chrY clusters
clusters %<>% filter(cluster.counts > 0, chr != "chrY", chr != "chrM") # 15,886

a375_screen_lfc_clusters = as.data.table(clusters)
save(a375_screen_lfc_clusters, file = "~/isoviz/data/a375_screen_lfc_clusters.rda")

Raw data -> Data for package
Gene/transcript name conversion file
Others might want to do this for a different gtf version
```{r}
# could make this a function
# create gene name conversion file from gtf

gtf_convert = function(gtf_file){
  gtf = read_tsv(gtf_file, col_names = FALSE, skip = 5) %>% filter(X3 == "transcript")

  gtf$transcript_id = sapply(strsplit(as.character(gtf$X9), ";"), function(x) x[[2]])
  gtf$transcript_id = sapply(gtf$transcript_id, gsub, pattern = "transcript_id", replacement = "")
  gtf$transcript_id = sapply(as.character(gtf$transcript_id), gsub, pattern = '"', replacement = '')
  gtf$transcript_id = sapply(as.character(gtf$transcript_id), gsub, pattern = ' ', replacement = '')

@@

  gtf %<>% dplyr::select(gene_id, trans_id = transcript_id, gene_name, transcript_name, gene_type, transcript_type) %>%
    distinct()

}

output = gtf_convert("~/isoviz/data-raw/gencode.v41.basic.annotation.gtf")
write_tsv(output, "~/isoviz/data-raw/gencode_v41_gene-transcript-convert.txt", col_names = TRUE)

@@

Add this to original genomedata.R script
```{r}
# modify psl file data to include other gene and transcript names
# currently using gencode basic anntotation file
convert = read_tsv("~/isoviz/data-raw/gencode_v41_gene-transcript-convert.txt", col_names = TRUE)

@@

Write function to generate intron coords from exon coords- this is currently slow (~30 minutes), can we make the loop more efficient?
```{r}
# make psl with intron coords instead of exon
# adapted from genomedata.R

make_intron_coords = function(psl){

  mydataset <- fread(psl)

@@

  # this is the part that is slow
  for(id in transcript_ids){
    data = mydataset %>% filter(trans_id == id)
    new_ends = data$blockstarts[-1] + 1
    len = nrow(data)
    new_starts = data$blockends[-len]
    n = len-1
    trans_id = c(trans_id, rep(id, n))
    intron_starts = c(intron_starts, new_starts)
    intron_ends = c(intron_ends, new_ends)
  }

@@

}

iso_intron_data = make_intron_coords("~/isoviz/data-raw/gencode.v41.basic.annotation.psl")
save(iso_intron_data, file = "~/isoviz/data/iso_intron_data.rda")

@@

Function: Map junction to isoform
```{r}
load(file = "~/isoviz/data/iso_intron_data.rda")
load(file = "~/isoviz/data/a375_screen_lfc_clusters.rda")

@@

  # filter for gene, for now, also filter for protein coding transcripts- can set this as a parameter later
  gene_iso_data = iso_intron_data %>% filter(grepl(gene, gene_id), transcript_type == "protein_coding") # 17 for NFYA

  # join iso data with leafcutter data and mark any unannotated junctions
  const_junc_list = a375_screen_lfc_clusters %>% 
    group_by(cluster.n) %>% 
    tally() %>% filter(n == 1) %>% ungroup()
  a375_screen_lfc_clusters %<>% mutate(junc.type = ifelse(cluster.n %in% const_junc_list$cluster.n, "Common", "Unique"))
  gene_cluster = gene_iso_data %>% 
    left_join(a375_screen_lfc_clusters, by = c("chr", "intron_starts" = "start", "intron_ends" = "end")) %>%
    filter(!is.na(cluster.n))

  gene = unique(gene_cluster$gene_id)
  name = unique(gene_cluster$gene_name)
  gene_strand = unique(gene_cluster$strand)
  cluster_ids = unique(gene_cluster$cluster.n)

  output = a375_screen_lfc_clusters %>% filter(cluster.n %in% cluster_ids) %>% group_by(cluster.n) %>% mutate(junc.per.cluster = n()) %>% 
    left_join(gene_iso_data, by = c("chr", "start" = "intron_starts", "end" = "intron_ends")) %>% 
    mutate(junc.usage = round((junc.counts/cluster.counts)*100, digits = 0), 
           transcript_name = ifelse(is.na(transcript_name), paste0(cluster.n, "_Unannotated"), transcript_name),
           trans_id = ifelse(is.na(trans_id), paste0(cluster.n, "_Unannotated"), trans_id),
           gene_id = ifelse(is.na(gene_id), gene, gene_id),
           gene_name = ifelse(is.na(gene_name), name, gene_name),
           strand = ifelse(is.na(strand), gene_strand, strand))

@@

to_plot = map_junction("ENSG00000001167.15") # NFYA
to_plot = map_junction("ENSG00000100320") # RBFOX2, alt start is missing because it's intron doesnt overlap another and counts are only 8 so included in consitutive either

# can we ask if all junctions of an isoform are represented to filter isoforms?
# can we create pseudo regtools output based on all possible junctions and run that through leafcutter to get true common versus unique

@@

```{r}
load(file='~/isoviz/data/iso_data_complete.rda')
df = to_plot

# plot isoform plus leafcutter clusters
plot_junction_to_isoform = function(df, include_common_juncs = FALSE){

  # I am copying the plot_isoforms function here, figure out how to integrate these later
  gene = unique(df$gene_id)
  gene_trans =which(str_detect(iso_data_complete$gene_id, gene))
  gene_data = iso_data_complete[gene_trans,]

  # filter out common junctions for plotting (default)
  if(include_common_juncs == FALSE){
    df %<>% filter(junc.type == "Unique")
  }

  # get info
  name = unique(df$gene_name)
  transcripts = unique(df$transcript_name)

@@

  to_plot_ordered = gene_data %>% full_join(order_plot, by = c("transcript_name")) %>%
    arrange(trans_order, blockstarts) 
  to_plot_ordered = to_plot_ordered %>% mutate(seg_end = end)

  # make isoform plot
  p1 = ggplot() +
    geom_segment(aes(x = to_plot_ordered$start, y = to_plot_ordered$trans_order,
                     xend = to_plot_ordered$seg_end, yend = to_plot_ordered$trans_order)) +
    geom_rect(data = to_plot_ordered, mapping = aes(xmin = blockstarts, xmax = blockends, ymin = trans_order - 0.3, ymax = trans_order + 0.3)) +
    xlim(min_start, max_end) + ylim(0,y_max) +
    geom_text(aes(x = Inf, y = to_plot_ordered$trans_order, hjust = 0, label = to_plot_ordered$transcript_name), size = 3) +
    theme_bw() + ylab("") + xlab("") +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(), 
          axis.text.y=element_blank(), axis.ticks.y=element_blank(), legend.position = "top", 
          plot.title = element_text(hjust = 0.5), plot.margin = margin(0.1,1,0,0.1, "in")) +
    ggtitle(paste0(name, " Junction to Isoform Map (" , to_plot_ordered$strand[1], " strand )")) + 
    coord_cartesian(xlim = c(min_start, max_end), clip = 'off')

  # new code to plot leafcutter cluster info underneath isoform level data
  # aggregate isoform information
  text = df %>% separate(col = transcript_name, into = c("name", "transcript")) %>% group_by(cluster.n, start, end) %>% 
    arrange(cluster.n, start, end, transcript) %>%
    mutate(isoforms = paste0(transcript, collapse = ",")) %>%
    select(1:5, junc.usage, isoforms) %>% distinct() %>% ungroup() %>% 
    mutate(text_plot = paste0(junc.counts, " reads (", junc.usage, "%) : ", isoforms))

  clusters = unique(df$cluster.n)
  introns = df %>% select(-gene_id, -trans_id, -gene_name, -transcript_name) %>% 
    distinct() %>% arrange(cluster.n, start)
  introns$junc.order = 1:nrow(introns)
  introns %<>% left_join(text)
  p2 = ggplot() + 
    geom_rect(data = introns, mapping = aes(xmin = start, xmax = end, ymin = junc.order - 0.2, ymax = junc.order + 0.2, fill = cluster.n)) + 
    geom_text(data = introns, aes(x = end, y = junc.order, label = text_plot), hjust = 0, size = 3) +
    xlim(min_start, max_end) + theme_bw() + 
    xlab("Hg38 Genomic Position (bp)") + ylab("") + 
    theme(axis.text.y=element_blank(), axis.ticks.y=element_blank(), legend.position = "none", plot.margin = margin(0,1,0,0.1, "in")) +
    coord_cartesian(xlim = c(min_start, max_end))

  # plot dimenstions based on number of isoforms and junctions
  t = (n + nrow(introns))
  h = t/2
  p1_p2 = plot_grid(p1, p2, ncol = 1, align = "v", rel_heights = c(n*0.8, nrow(introns)))
  #ggsave("~/isoviz/plots/rbfox2_isoforms.pdf", width = 8, height = h)
  return(p1_p2)
}

plot_junction_to_isoform(to_plot)
ggsave("~/isoviz/plots/rbfox2_isoforms.pdf", width = 8, height = 8.5)
```

```{r}
to_plot = map_junction("ENSG00000001167.15") # NFYA
plot_junction_to_isoform(to_plot)
ggsave("~/isoviz/plots/nfya_isoforms.pdf", width = 8, height = 4)

to_plot = map_junction("ENSG00000100320") # RBFOX2, alt start is missing because it's intron doesnt overlap another and counts are only 8 so included in consitutive either
plot_junction_to_isoform(to_plot)
```
